---
title: "Differential expression and GO analysis"
output:
  pdf_document:
    toc: true
    number_sections: true
params:
  quant_method: "hisat2"        # hisat2 | kallisto
  config_path: "../config"
  utils_path: "../scripts/R/utils.R"
  output_dir: "results"
---

```{r setup, include=FALSE}
library(yaml)
library(limma)
library(edgeR)
library(DESeq2)
library(sva)
library(topGO)
library(biomaRt)
library(ggplot2)
library(gplots)
library(RColorBrewer)
library(VennDiagram)
library(gridExtra)
library(statmod)

config <- read_yaml(file.path(params$config_path, "config.yaml"))

source(params$utils_path)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r data_loading}

Description <- read.csv(file.path(params$config_path, "Description.csv"))

if (max(Description$Repetition) < 2) {
  stop("Too few repetitions (min 2 required)")
}

if (params$quant_method == "hisat2") {

  base_path <- file.path(config$project$output_dir,
                         config$project$name,
                         "genome")

  count_matrix <- read.csv(
    file.path(base_path, "gene_count_matrix.csv"),
    row.names = 1
  )

} else if (params$quant_method == "kallisto") {

  base_path <- file.path(config$project$output_dir,
                         config$project$name,
                         "trans/kallisto")

  data_list <- lapply(Description$Group, function(name) {
    read.table(
      file.path(base_path, name, "abundance.tsv"),
      header = TRUE,
      row.names = 1
    )$est_counts
  })

  count_matrix <- do.call(cbind, data_list)
  rownames(count_matrix) <- rownames(
    read.table(file.path(base_path,
                         Description$Group[1],
                         "abundance.tsv"),
               header = TRUE,
               row.names = 1)
  )
}

dataset_size <- nrow(Description)

Description$Name <- paste(
  "B",
  Description$Batch,
  "_",
  Description$Condition,
  sep=""
)

colnames(count_matrix) <- Description$Name
count_matrix <- count_matrix[, order(colnames(count_matrix))]
```
```{r filtering}

th1 <- 5
th2 <- 2

n1 <- round(ncol(count_matrix)/3)
n2 <- n1*2

filter1 <- apply(count_matrix,1,function(x) length(x[x>th1])>=n1)
filter2 <- apply(count_matrix,1,function(x) length(x[x>th2])>=n2)

filter <- filter1 & filter2

counts_filtered <- count_matrix[filter,]
counts_log <- log2(counts_filtered+0.5)
```

# Heatmap raw data

```{r,warning=FALSE,message=FALSE,fig.align='left'}
IQR <- apply(counts_filtered,1,IQR)
topVar <- order(IQR,decreasing=TRUE)[1:100]

heatmap.2(as.matrix(counts_log[topVar,]),
          scale="row",
          trace="none",
          density="none",
          labRow="",
          dendrogram="column",
          key=FALSE)
```

```{r design_setup}
samples <- sort(paste("B",Description$Batch,"_",Description$Condition,sep=""))
names_col <- unique(samples)
group_size <- length(samples)/length(names_col)

no_of_batches <- length(unique(Description$Batch))
size_of_batches <- table(Description$Batch)

qual_col_pals <- brewer.pal.info[brewer.pal.info$category=="qual",]
col_vector <- unlist(mapply(brewer.pal,
                            qual_col_pals$maxcolors,
                            rownames(qual_col_pals)))
```

# PCA raw data

```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=7,fig.width=7}
Plot_PCA(counts_log,
         samples,
         names_col,
         group_size,
         no_of_batches,
         size_of_batches,
         col_vector,
         "PCA before SVA")
```

# Violinplot raw data

```{r,warning=FALSE,message=FALSE,fig.height = 5}
Plot_violin(counts_filtered, samples, group_size)
```

```{r limma_sva}
#############################
# LIMMA + VOOM + ROBUST SVA 
#############################

dge <- DGEList(counts = counts_filtered, group = samples)
dge <- calcNormFactors(dge)

## Design matrices
mod1 <- model.matrix(~0 + as.factor(samples))
mod0 <- model.matrix(~1, data = as.factor(samples))

colnames(mod1) <- levels(factor(samples))
rownames(mod1) <- colnames(counts_filtered)

## Voom transform
v <- voom(dge, mod1, plot = FALSE)

to_sva <- 2^v$E

#############################
# ESTIMATE NUMBER OF SVs
#############################

nsv <- num.sv(as.matrix(counts_filtered), mod1, method = "leek")

an.error.occured <- TRUE

while (an.error.occured == TRUE) {

  svseq <- try(
    svaseq(to_sva, mod1, mod0, const = 0, n.sv = nsv),
    TRUE
  )

  if (isTRUE(class(svseq) == "try-error")) {
    nsv <- nsv - 1
    if (nsv <= 0) stop("SVA failed: nsv dropped to 0.")
    next
  } else {
    an.error.occured <- FALSE
  }
}

fsvaobj <- fsva(v$E, mod1, svseq, newdat = NULL)
```

# PCA with all factors removed `r nsv``

```{r,warning=FALSE,message=FALSE,fig.height = 7, fig.width=7}
Plot_PCA(fsvaobj$db,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA with all factors removed (",nsv,")"))
```


```{r,warning=FALSE,message=FALSE}

############################################################
# BUILD FULL MODEL INCLUDING SVs
############################################################

mmod1 <- cbind(mod1, svseq$sv)
colnames(mmod1) <- c(
  levels(factor(samples)),
  paste("X", 1:nsv, sep = "")
)

v2 <- voom(dge, mmod1, plot = FALSE)

############################################################
# SAFE LIMMA FIT LOOP 
############################################################

an.error.occured <- TRUE

while (an.error.occured == TRUE) {

  ok <- TRUE

  fit <- tryCatch(
    lmFit(v2, mmod1, method = "robust", maxit = 9999),
    warning = function(w) {
      ok <<- FALSE
    }
  )

  if (!ok) {
    nsv <- nsv - 1
    if (nsv <= 0) stop("Limma fit failed: nsv dropped to 0.")

    svseq <- svaseq(to_sva, mod1, mod0, const = 0, n.sv = nsv)

    mmod1 <- cbind(mod1, svseq$sv)
    colnames(mmod1) <- c(
      levels(factor(samples)),
      paste("X", 1:nsv, sep = "")
    )

    v2 <- voom(dge, mmod1, plot = FALSE)

    next
  } else {
    an.error.occured <- FALSE
  }
}

fsvaobj <- fsva(v$E, mod1, svseq, newdat = NULL)
```

# Final PCA with `r nsv` factors removed

```{r,warning=FALSE,message=FALSE,fig.height = 7}
Plot_PCA(fsvaobj$db,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA with ",nsv," factors removed"))

```

```{r,warning=FALSE,message=FALSE}

contrast_string <- config$CONTRAST
P_val <- config$P_VAL

cm <- makeContrasts(contrast_string, levels = mmod1)

cf <- contrasts.fit(fit, cm)
fe <- eBayes(cf)

tt <- topTable(fe, num = Inf)

pr <- 1 - propTrueNull(tt$P.Value, method = "convest")
cfe <- eBayes(cf, proportion = pr)

res_limma <- topTable(
  cfe,
  number = Inf,
  adjust.method = "BH",
  sort.by = "none"
)

res_limma2 <- res_limma[res_limma$adj.P.Val < P_val, ]
```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_limma2)[1] ``

# Section 2 -EdgeR

```{r edger}
y <- DGEList(counts=counts_filtered, group=samples)
keep <- filterByExpr(y)
y <- y[keep,,keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

y <- estimateDisp(y,mmod1,robust=TRUE)
fit_edge <- glmQLFit(y,mmod1,robust=TRUE)

qlf <- glmQLFTest(fit_edge,contrast=cm)

res_edgeR <- topTags(qlf,Inf)$table
res_edgeR2 <- res_edgeR[res_edgeR$FDR<P_val,]
```


# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_edgeR2)[1] ``


# Section 3 -DeSeq
 
```{r deseq2}
coldata <- data.frame(condition=samples)

dds <- DESeqDataSetFromMatrix(countData=round(counts_filtered),
                              colData=coldata,
                              design=~condition)

dds <- DESeq(dds)

res_deseq <- results(dds,
                     contrast=c("condition",
                                config$GROUP1,
                                config$GROUP2))

res_deseq2 <- res_deseq[res_deseq$padj<P_val,]
```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_deseq2)[1] `

# Venn diagram for all three methods

```{r venn}
venn.diagram(
  x=list(
    Limma=rownames(res_limma2),
    edgeR=rownames(res_edgeR2),
    DESeq2=rownames(res_deseq2)
  ),
  filename=NULL,
  fill=c("yellow","lightgreen","orchid")
)
```

# Barplots for top 10 terms for given GO term


```r
```{r go}

methods <- c("limma", "edgeR", "DeSeq")
ontologies <- c("BP", "MF", "CC")

GO_results <- list()

for (method in methods) {

  res_obj <- get(paste0("res_", method, "2"))

  if (nrow(res_obj) == 0) next

  geneList <- res_obj$adj.P.Val
  names(geneList) <- rownames(res_obj)

  for (ont in ontologies) {

    GOdata <- new("topGOdata",
                  ontology = ont,
                  allGenes = geneList,
                  geneSel = topDiffGenes,
                  nodeSize = 10,
                  annot = annFUN.org,
                  mapping = config$GO_MAPPING,
                  ID = "ensembl")

    resultFisher <- runTest(GOdata,
                            algorithm = "parentchild",
                            statistic = "fisher")

    used <- usedGO(GOdata)

    GO.allRes <- GenTable(GOdata,
                          parentchildFisher = resultFisher,
                          orderBy = "parentchildFisher",
                          ranksOf = "parentchildFisher",
                          topNodes = length(used))

    colnames(GO.allRes)[6] <- "adj.P.Val"

    GO.allRes2 <- GO.allRes[
      as.numeric(GO.allRes$adj.P.Val) < config$TH_GO,
    ]

    ############################################################
    # BARPLOT TOP 10
    ############################################################

    if (nrow(GO.allRes2) > 0) {

      b <- min(10, nrow(GO.allRes2))

      # handle "< 1e-30"
      GO.allRes2$adj.P.Val[GO.allRes2$adj.P.Val == "< 1e-30"] <- 1e-30
      GO.allRes2$adj.P.Val <- as.numeric(GO.allRes2$adj.P.Val)

      par(mai = c(1, 3, 1, 1))

      barplot(
        abs(log(GO.allRes2$adj.P.Val[b:1])),
        horiz = TRUE,
        names.arg = GO.allRes2$Term[b:1],
        main = paste("Top", b, ont, "for", method),
        xlab = "abs(log(p-value))",
        las = 1,
        col = "darkblue",
        cex.names = 0.7
      )
    }

    ############################################################
    # SAVE
    ############################################################

    write.csv(GO.allRes,
              paste0("Results_", method, "_", ont, ".csv"),
              row.names = FALSE)

    write.csv(GO.allRes2,
              paste0("Results_", method, "_", ont,
                     "_", config$TH_GO, ".csv"),
              row.names = FALSE)

    # store significant GO IDs for Venn
    GO_results[[ont]][[method]] <- GO.allRes2$GO.ID
  }
}
```

# Venn diagrams for topGO

```{r,warning=FALSE,message=FALSE}

for (ont in ontologies) {

  if (is.null(GO_results[[ont]])) next

  # remove empty method entries
  venn_list <- GO_results[[ont]]
  venn_list <- venn_list[sapply(venn_list, length) > 0]

  if (length(venn_list) < 2) {
    message("Not enough results for Venn in ", ont)
    next
  }

  ven <- venn.diagram(
    x = venn_list,
    filename = NULL,
    fill = brewer.pal(length(venn_list), "Set2"),
    main = paste("GO", ont)
  )

  grid::grid.newpage()
  grid::grid.draw(ven)
}

```