---
title: "Differential expression and GO analysis"
output:
  pdf_document:
    toc: true
    number_sections: true
params:
  quant_method: "hisat2"
  config_path: "config"
  samples_path: "Test_data/samples.tsv"
  utils_path: "scripts/utils.R"
---

```{r setup, include=FALSE}
library(yaml)
library(limma)
library(edgeR)
library(DESeq2)
library(sva)
library(topGO)
library(biomaRt)
library(ggplot2)
library(gplots)
library(RColorBrewer)
library(VennDiagram)
library(gridExtra)
library(statmod)
library(GenomicFeatures)

#For debug
#params <- list(
#  quant_method = "hisat2",
# config_path = "config",
#  samples_path = "Test_data/samples.tsv",
#  utils_path = "scripts/utils.R"
#)
#For debug
config <- read_yaml(file.path(params$config_path, "config.yaml"))

source(params$utils_path)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
output_dir <- file.path(
  config$project$output_dir,
  config$project$name,
  "dea"
)

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r data_loading}

Description <- read.table(
  params$samples_path,
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE
)

if (max(Description$Repetition) < 2) {
  stop("Too few repetitions (min 2 required)")
}

if (params$quant_method == "hisat2") {

  base_path <- file.path(config$project$output_dir,
                         config$project$name,
                         "genome")

  count_matrix <- read.csv(
    file.path(base_path, "gene_count_matrix.csv"),
    row.names = 1
  )

} else if (params$quant_method == "kallisto") {

  base_path <- file.path(config$project$output_dir,
                         config$project$name,
                         "trans/kallisto")

  tx_list <- lapply(Description$sample, function(name) {
    tab <- read.table(
      file.path(base_path, name, "abundance.tsv"),
      header = TRUE,
      stringsAsFactors = FALSE
    )
    tab[, c("target_id", "est_counts")]
  })

  tx_ids <- tx_list[[1]]$target_id

  count_matrix_tx <- do.call(
    cbind,
    lapply(tx_list, function(x) x$est_counts)
  )

  rownames(count_matrix_tx) <- tx_ids
  colnames(count_matrix_tx) <- Description$sample

  gtf_path <- file.path("Test_data", "annotation.gtf")

  txdb <- makeTxDbFromGFF(gtf_path)

  tx2gene <- select(txdb,
                    keys = keys(txdb, keytype = "TXNAME"),
                    keytype = "TXNAME",
                    columns = "GENEID")

  colnames(tx2gene) <- c("TXNAME", "GENEID")

  tx_df <- as.data.frame(count_matrix_tx)
  tx_df$TXNAME <- rownames(tx_df)

  tx_df <- merge(tx_df, tx2gene, by = "TXNAME")

  count_matrix_gene <- aggregate(
    . ~ GENEID,
    data = tx_df[, -1],
    FUN = sum
  )

  rownames(count_matrix_gene) <- count_matrix_gene$GENEID
  count_matrix_gene$GENEID <- NULL

  count_matrix <- as.matrix(count_matrix_gene)
}

dataset_size <- nrow(Description)

Description$Name <- paste0(
  "B", Description$Batch,
  "_", Description$Condition,
  "_R", Description$Repetition
)

colnames(count_matrix) <- Description$Name
count_matrix <- count_matrix[, order(colnames(count_matrix))]
```
```{r filtering}

th1 <- 5
th2 <- 2

n1 <- round(ncol(count_matrix)/3)
n2 <- n1*2

filter1 <- apply(count_matrix,1,function(x) length(x[x>th1])>=n1)
filter2 <- apply(count_matrix,1,function(x) length(x[x>th2])>=n2)

filter <- filter1 & filter2

counts_filtered <- count_matrix[filter,]
counts_log <- log2(counts_filtered+0.5)
```

# Heatmap raw data

```{r,warning=FALSE,message=FALSE,fig.align='left'}
IQR <- apply(counts_filtered,1,IQR)
n_top <- min(100, nrow(counts_log))
topVar <- order(IQR, decreasing=TRUE)[1:n_top]

heatmap.2(as.matrix(counts_log[topVar,]),
          scale="row",
          trace="none",
          density="none",
          labRow="",
          dendrogram="column",
          key=FALSE)
```

```{r design_setup}
samples <- sort(paste("B",Description$Batch,"_",Description$Condition,sep=""))
names_col <- unique(samples)
group_size <- length(samples)/length(names_col)

no_of_batches <- length(unique(Description$Batch))
size_of_batches <- table(Description$Batch)

qual_col_pals <- brewer.pal.info[brewer.pal.info$category=="qual",]
col_vector <- unlist(mapply(brewer.pal,
                            qual_col_pals$maxcolors,
                            rownames(qual_col_pals)))
```

# PCA raw data

```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=7,fig.width=7}
Plot_PCA(counts_log,
         samples,
         names_col,
         group_size,
         no_of_batches,
         size_of_batches,
         col_vector,
         "PCA before SVA")
```

# Violinplot raw data

```{r,warning=FALSE,message=FALSE,fig.height = 5}
Plot_violin(counts_filtered, samples, group_size)
```

```{r limma_sva}
#############################
# LIMMA + VOOM 
#############################

dge <- DGEList(counts = counts_filtered, group = samples)
dge <- calcNormFactors(dge)

mod1 <- model.matrix(~0 + as.factor(samples))
mod0 <- model.matrix(~1, data = as.factor(samples))
colnames(mod1) <- levels(factor(samples))
rownames(mod1) <- colnames(counts_filtered)

v <- voom(dge, mod1, plot = FALSE)
to_sva <- 2^v$E

#############################
# SVA — opcjonalne
#############################

sva_enabled <- isTRUE(config$sva$enabled)
sva_success <- FALSE
nsv <- 0

if (sva_enabled) {
  
  nsv <- tryCatch(
    num.sv(as.matrix(counts_filtered), mod1, method = "leek"),
    error   = function(e) { message("num.sv failed: ", e$message); 0 },
    warning = function(w) { message("num.sv warning: ", w$message); 0 }
  )
  
  if (nsv > 0) {
    
    while (nsv > 0) {
      svseq <- tryCatch(
        svaseq(to_sva, mod1, mod0, const = 0, n.sv = nsv),
        error   = function(e) NULL,
        warning = function(w) NULL
      )
      
      if (!is.null(svseq)) {
        sva_success <- TRUE
        break
      }
      
      nsv <- nsv - 1
      message("SVA failed, retrying with nsv=", nsv)
    }
  }
  
  if (!sva_success) {
    message("SVA disabled: fallback to no batch correction")
  }
}

#############################
# FALLBACK gdy SVA nie działa
#############################

if (!sva_success) {
  nsv   <- 0
  mmod1 <- mod1
  v2    <- v
  # fsvaobj.db = surowe v$E (bez korekcji)
  fsvaobj <- list(db = v$E)
} else {
  fsvaobj <- fsva(v$E, mod1, svseq, newdat = NULL)
  mmod1   <- cbind(mod1, svseq$sv)
  colnames(mmod1) <- c(
    levels(factor(samples)),
    paste("X", 1:nsv, sep = "")
  )
  v2 <- voom(dge, mmod1, plot = FALSE)
}
```

```{r pca_title_1, results='asis'}
cat(paste0("# PCA — czynniki usunięte: ", nsv, "\n"))

Plot_PCA(fsvaobj$db, samples, names_col, group_size,
         no_of_batches, size_of_batches, col_vector,
         paste("PCA with", nsv, "factors removed"))

```



```{r limma_fit}
############################################################
# SAFE LIMMA FIT LOOP 
############################################################

an.error.occured <- TRUE

while (an.error.occured == TRUE) {
  
  ok <- TRUE
  
  fit <- tryCatch(
    lmFit(v2, mmod1, method = "robust", maxit = 9999),
    warning = function(w) { ok <<- FALSE }
  )
  
  if (!ok) {
    
    if (!sva_success || nsv <= 0) {
      # brak SVA — nie możemy redukować, prosta metoda
      message("Robust fit failed, fallback to standard lmFit")
      fit <- lmFit(v2, mmod1)
      an.error.occured <- FALSE
      next
    }
    
    nsv <- nsv - 1
    message("Limma fit failed, retrying with nsv=", nsv)
    
    if (nsv == 0) {
      mmod1   <- mod1
      v2      <- v
      fsvaobj <- list(db = v$E)
    } else {
      svseq <- svaseq(to_sva, mod1, mod0, const = 0, n.sv = nsv)
      mmod1 <- cbind(mod1, svseq$sv)
      colnames(mmod1) <- c(
        levels(factor(samples)),
        paste("X", 1:nsv, sep = "")
      )
      v2      <- voom(dge, mmod1, plot = FALSE)
      fsvaobj <- fsva(v$E, mod1, svseq, newdat = NULL)
    }
    
  } else {
    an.error.occured <- FALSE
  }
}

```

# Final PCA with `r nsv` factors removed

```{r,warning=FALSE,message=FALSE,fig.height = 7}
Plot_PCA(fsvaobj$db,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA with ",nsv," factors removed"))

```

```{r,warning=FALSE,message=FALSE}

contrast_string <- config$dge$contrast
P_val <- config$dge$p_value

cm <- makeContrasts(contrast_string, levels = mmod1)

cf <- contrasts.fit(fit, cm)
fe <- eBayes(cf)

tt <- topTable(fe, num = Inf)

pr <- 1 - propTrueNull(tt$P.Value, method = "convest")
cfe <- eBayes(cf, proportion = pr)

res_limma <- topTable(
  cfe,
  number = Inf,
  adjust.method = "BH",
  sort.by = "none"
)

res_limma2 <- res_limma[res_limma$adj.P.Val < P_val, ]
```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_limma2)[1] ``

# Section 2 -EdgeR

```{r edger}
y <- DGEList(counts=counts_filtered, group=samples)
keep <- filterByExpr(y)
y <- y[keep,,keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

y <- estimateDisp(y,mmod1,robust=FALSE)
fit_edge <- glmQLFit(y,mmod1,robust=TRUE)

qlf <- glmQLFTest(fit_edge,contrast=cm)

res_edgeR <- topTags(qlf,Inf)$table
res_edgeR2 <- res_edgeR[res_edgeR$FDR<P_val,]
```


# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_edgeR2)[1] ``


# Section 3 -DeSeq
 
```{r deseq2}
coldata <- data.frame(condition = samples)

dds <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData   = coldata,
  design    = ~condition
)

dds <- estimateSizeFactors(dds)

dds <- tryCatch({

  estimateDispersions(dds)

}, error = function(e) {

  message(
    "Standard dispersion fitting failed, ",
    "using gene-wise estimates as final: ",
    e$message
  )
  dds <- estimateDispersionsGeneEst(dds)
  dispersions(dds) <- mcols(dds)$dispGeneEst
  dds

})

dds <- nbinomWaldTest(dds)

batch  <- unique(Description$Batch)[1]
g1 <- paste0("B", batch, "_", config$dge$group1)
g2 <- paste0("B", batch, "_", config$dge$group2)

res_deseq <- results(
  dds,
  contrast = c("condition", g1, g2)
)

res_deseq2 <- as.data.frame(
  res_deseq[!is.na(res_deseq$padj) & res_deseq$padj < P_val, ]
)

```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_deseq2)[1] `

# Venn diagram for all three methods

```{r venn}
ven<-venn.diagram(
  x=list(
    Limma=rownames(res_limma2),
    edgeR=rownames(res_edgeR2),
    DESeq2=rownames(res_deseq2)
  ),
  filename=NULL,
  fill=c("yellow","lightgreen","orchid"),
disable.logging = TRUE
)
grid::grid.newpage()
grid::grid.draw(ven)
```

# Barplots for top 10 terms for given GO term


```{r go}

methods <- c("limma", "edgeR", "deseq")
ontologies <- c("BP", "MF", "CC")

GO_results <- list()

for (method in methods) {

  res_obj <- get(paste0("res_", method))

  if (nrow(res_obj) == 0) next

  get_pval_col <- function(res_obj) {
  if ("adj.P.Val" %in% colnames(res_obj)) return(res_obj$adj.P.Val)
  if ("FDR"       %in% colnames(res_obj)) return(res_obj$FDR)
  if ("padj"      %in% colnames(res_obj)) return(res_obj$padj)
  }
  geneList <- get_pval_col(res_obj)
  gene_ids <- sub("\\|.*", "", rownames(res_obj))
  names(geneList) <- gene_ids

  for (ont in ontologies) {

    GOdata <- new("topGOdata",
                  ontology = ont,
                  allGenes = geneList,
                  geneSel = topDiffGenes,
                  nodeSize = 10,
                  annot = annFUN.org,
                  mapping = config$dge$go_mapping,
                  ID = "ensembl")

    resultFisher <- runTest(GOdata,
                            algorithm = "parentchild",
                            statistic = "fisher")

    used <- usedGO(GOdata)

    GO.allRes <- GenTable(GOdata,
                          parentchildFisher = resultFisher,
                          orderBy = "parentchildFisher",
                          ranksOf = "parentchildFisher",
                          topNodes = length(used))

    colnames(GO.allRes)[6] <- "adj.P.Val"

    GO.allRes2 <- GO.allRes[
      as.numeric(GO.allRes$adj.P.Val) < config$dge$go_threshold,
    ]

    ############################################################
    # BARPLOT TOP 10
    ############################################################

    if (nrow(GO.allRes2) > 0) {

      b <- min(10, nrow(GO.allRes2))

      # handle "< 1e-30"
      GO.allRes2$adj.P.Val[GO.allRes2$adj.P.Val == "< 1e-30"] <- 1e-30
      GO.allRes2$adj.P.Val <- as.numeric(GO.allRes2$adj.P.Val)

      par(mai = c(1, 3, 1, 1))

      barplot(
        abs(log(GO.allRes2$adj.P.Val[b:1])),
        horiz = TRUE,
        names.arg = GO.allRes2$Term[b:1],
        main = paste("Top", b, ont, "for", method),
        xlab = "abs(log(p-value))",
        las = 1,
        col = "darkblue",
        cex.names = 0.7
      )
    }

    ############################################################
    # SAVE
    ############################################################

    write.csv(GO.allRes,
              paste0(output_dir,"/Results_", method, "_", ont, ".csv"),
              row.names = FALSE)

    write.csv(GO.allRes2,
              paste0(output_dir,"/Results_", method, "_", ont,
                     "_", config$dge$go_threshold, ".csv"),
              row.names = FALSE)

    # store significant GO IDs for Venn
    GO_results[[ont]][[method]] <- GO.allRes2$GO.ID
  }
}
```

# Venn diagrams for topGO

```{r,warning=FALSE,message=FALSE}

for (ont in ontologies) {

  if (is.null(GO_results[[ont]])) next

  # remove empty method entries
  venn_list <- GO_results[[ont]]
  venn_list <- venn_list[sapply(venn_list, length) > 0]

  if (length(venn_list) < 2) {
    message("Not enough results for Venn in ", ont)
    next
  }

  ven <- venn.diagram(
    x = venn_list,
    filename = NULL,
    fill = brewer.pal(length(venn_list), "Set2"),
    main = paste("GO", ont),
    disable.logging = TRUE
  )

  grid::grid.newpage()
  grid::grid.draw(ven)
}

write.csv(res_limma,
          file.path(output_dir, "DEG_limma_all.csv"))

write.csv(res_limma2,
          file.path(output_dir, "DEG_limma_sig.csv"))

write.csv(res_edgeR,
          file.path(output_dir, "DEG_edgeR_all.csv"))

write.csv(res_edgeR2,
          file.path(output_dir, "DEG_edgeR_sig.csv"))

write.csv(as.data.frame(res_deseq),
          file.path(output_dir, "DEG_deseq_all.csv"))

write.csv(res_deseq2,
          file.path(output_dir, "DEG_deseq_sig.csv"))
```