---
title: "Differential expresion and GO terms analysis with Kallisto"
output: pdf_document
params:
  inpath: "../configs"
---

```{r,warning=FALSE,message=FALSE,include=FALSE}

packages<-c("knitr","RColorBrewer","VennDiagram","veccompare","ggplot2","gplots","yaml","gridExtra","statmod")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE,repos = "http://cran.us.r-project.org")
      library(x, character.only = TRUE)
    }
  }
)
if (!require("BiocManager", quietly = TRUE))
   { install.packages("BiocManager",repos = "http://cran.us.r-project.org")
BiocManager::install()}
listOfBiocPackages<-c("topGO","org.Mm.eg.db","biomaRt","limma","DESeq2","edgeR","sva")
notInstalled <- which( !listOfBiocPackages %in% rownames(installed.packages()) )
## check there's still something left to install
if( length(notInstalled) ) {
    BiocManager::install(listOfBiocPackages[ notInstalled ])
}
lapply(listOfBiocPackages,library, character.only=TRUE)
lapply(packages,require, character.only=TRUE)

path_configs<-params$inpath
#opts_knit$set(root.dir = path)
knitr::opts_chunk$set(echo=FALSE)
config<-read_yaml(paste(path_configs,'/config.yaml',sep=''),fileEncoding = "UTF-8")
kallisto_path<-paste(config$FINALOUTPUT,config$PROJECT,'trans/kallisto',sep='/')

```

```{r,warning=FALSE,message=FALSE,include=FALSE}
###########
#Functions
###########

Plot_PCA<-function(data,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,title){

    symbols<-c()
    for (i in 1:no_of_batches){
        symbols<-c(symbols,c(rep(i,each=size_of_batches[i])))
        }

    pc<-prcomp(t(data))
    pc1<-round((pc$sdev[1])^2 / sum(pc$sdev^2),digits=2)
    pc2<-round((pc$sdev[2])^2 / sum(pc$sdev^2),digits=2) 
    pc3<-round((pc$sdev[3])^2 / sum(pc$sdev^2),digits=2) 
    cols<-rep(sample(col_vector,length(names_col)),each=group_size)
    par(mar=c(5.1, 4.1, 4.1, 8.1), xpd=TRUE)
    plot( pc$x[,1:2],pch=symbols, cex=3,col=cols, main=title,xlab=paste('PC1=',pc1),ylab=paste('PC2=',pc2))
    #plot( pc$x[,2:3],pch=symbols, cex=3,col=cols, main=title,xlab=paste('PC2=',pc2),ylab=paste('PC3=',pc3))
    legend("topright", inset=c(-0.3,0),legend=names_col,pch=rep(unique(symbols),size_of_batches/group_size),col=unique(cols),cex=0.7)
}



Plot_violin<-function(data,samples, group_size){

    variable<-group_size-1
    for (i in seq(1,length(samples)-variable, by=group_size))
    {
        disp<-paste(samples[i],'_disp_raw',sep="")
        disp_v<-apply(data[,i:(i+variable)],1,function(x) (
            if (sum(x)>0){
                (sd(x, na.rm=TRUE)/mean(x, na.rm=TRUE))*100}
            else{200}
        ))
        assign(paste(disp),disp_v)
    }

    df_dispersion_raw<-NULL
    for (i in (1:length(names_col)))
    {
        df_dispersion_raw<-cbind(df_dispersion_raw,get(paste(names_col[i],'_disp_raw',sep='')))
    }
    colnames(df_dispersion_raw)<-paste(names_col,'_disp_raw',sep='')
    df_dispersion_raw<-as.data.frame(df_dispersion_raw)

    pr_below_10_raw<-c()
    pr_below_20_raw<-c()
    for (i in seq(1,length(df_dispersion_raw)))
    {
        pr_below_10_raw[i]<-(length(which(df_dispersion_raw[,i]<10))/dim(df_dispersion_raw)[1])*100
        pr_below_20_raw[i]<-(length(which(df_dispersion_raw[,i]<20))/dim(df_dispersion_raw)[1])*100
    }

    df.m <- reshape2::melt(df_dispersion_raw, id.vars = NULL)
    pl<-ggplot(df.m, aes(x = variable, y = value)) + geom_violin()
    pl + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
    
}
```

# Section 1 - Limma

```{r,warning=FALSE,message=FALSE,include=FALSE}

###################
#Data load
###################

## reading in kallisto
Description<-read.table(paste(path_configs,'/Description.csv',sep=''),header=TRUE,sep=',', stringsAsFactors=FALSE)
dataset_size<-dim(Description)[1]
for (i in (1:dataset_size))
{
  Description[i,7]<-paste('B',Description[i,5],'_',Description[i,3],'_',Description[i,4],sep="")
}
colnames(Description)[7]<-c("Name")


data.trans.counts <- c()
for (name in Description$Group)
{
  fin <- read.table(paste(kallisto_path,'/',name,'/abundance.tsv', sep=""), header=TRUE, row.names=1)
  data.trans.counts <- cbind(data.trans.counts, fin$est_counts)
}
rownames(data.trans.counts) <- rownames(fin)
colnames(data.trans.counts) <- Description$Name

kallisto.t <- data.trans.counts[,order(colnames(data.trans.counts))]

###################
#Data description
###################

no_of_batches<-length(unique(Description$Batch))
size_of_batches<-c()
for (i in (1:no_of_batches))
{
  size_of_batches<-cbind(size_of_batches,length(which(Description$Batch==i)))
}
if(sum(size_of_batches)!=dim(kallisto.t)[2]){
    stop('Sum of all batches sizes does not match input size')
    }
samples<-sort(paste('B',Description$Batch,'_',Description$Condition,sep=''))
names_col<-unique(samples)
group_size<-length(samples)/length(names_col)
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

######################
#Low values removal
######################

th1 <- 5
th2 <- 2
n1 <- round(dim(kallisto.t)[2]/3)
n2 <- n1*2
filter1 <- apply(kallisto.t, 1, function(x) length(x[x>th1])>=n1)
filter2 <- apply(kallisto.t, 1, function(x) length(x[x>th2])>=n2)
filter <- filter1&filter2

kallisto.t.filtered<-kallisto.t[filter,]
kallisto.t.filtered.log<-log2(kallisto.t.filtered+0.5)
 
```

# Heatmap raw data

```{r,warning=FALSE,message=FALSE,fig.align='left'}

IQR<- apply(kallisto.t.filtered,1,IQR,na.rm=TRUE)
topVar<-order(IQR, decreasing=TRUE)[1:100]
heatmap.2(as.matrix(kallisto.t.filtered.log[topVar,]),
          scale="row",
          hclust=function(x) hclust(x,method="average"),
          distfun=function(x) as.dist((1-cor(t(x)))/2),
          trace="none",
          main= 'Heatmap for 100 most varible samples\n raw data',
          density="none",
          labRow="",
          cexCol=0.5,dendrogram='column',key=FALSE)

```

# PCA raw data

```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=7,fig.width=7}

############
#PCA before
############
Plot_PCA(kallisto.t.filtered.log,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,"PCA before SVA")
#legend('topleft',legend=names_col,pch=rep(unique(symbols2),lwd=4,size_of_batches/group_size),col=unique(cols2),cex=0.7)


```
# Violinplot raw data

```{r,warning=FALSE,message=FALSE,fig.height = 5}
Plot_violin(kallisto.t.filtered,samples,group_size)
```

```{r,warning=FALSE,message=FALSE,include=FALSE}

########
##limma
########

dge <- DGEList(counts=kallisto.t.filtered, group=samples)
dge <- calcNormFactors(dge)
## voom transformation
mod1 <- model.matrix(~0+as.factor(samples))
mod0 <- model.matrix(~1,data=as.factor(samples))
colnames(mod1) <- levels(factor(samples))
rownames(mod1) <- colnames(kallisto.t.filtered)
v <- voom(dge,mod1,plot=FALSE)
to_sva <- 2^v$E
nsv<-num.sv(as.matrix(kallisto.t.filtered),mod1,method="leek") 
an.error.occured <- TRUE
while(an.error.occured==TRUE){
   svseq <- try(svaseq(to_sva,mod1,mod0,const=0,n.sv=nsv),TRUE)
   if(isTRUE(class(svseq)=="try-error")) { 
     nsv <-nsv-1
     next
      } 
   else { an.error.occured <- FALSE}
}
newV=NULL
fsvaobj<-fsva(v$E,mod1,svseq,newdat=NULL)

# names_col_new<-names_col
# names_col_new<-sub(paste("(B\\d_",GR1,")",sep=''),paste(GR1),names_col_new)
# names_col_new<-sub(paste("(B\\d_",GR2,")",sep=''),paste(GR2),names_col_new)
```

# PCA with all factors removed `r nsv``

```{r,warning=FALSE,message=FALSE,fig.height = 7, fig.width=7}
Plot_PCA(fsvaobj$db,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA with all factors removed (",nsv,")"))
```

```{r,warning=FALSE,message=FALSE,include=FALSE}
mmod1 <- cbind(mod1, svseq$sv)
colnames(mmod1) <- c(levels(factor(samples)), paste("X",1:nsv,sep=""))
v2 <- voom(dge,mmod1,plot=FALSE)
an.error.occured <- TRUE
while(an.error.occured==TRUE){
  ok<-TRUE
  fit<-tryCatch(lmFit(v2,mmod1,method='robust', maxit=9999),
         warning=function(w){

        ok<<-FALSE
         })
  if (!ok)
  {
    nsv <-nsv-1
    svseq <- svaseq(to_sva,mod1,mod0,const=0,n.sv=nsv)
    mmod1 <- cbind(mod1, svseq$sv)
    colnames(mmod1) <- c(levels(factor(samples)), paste("X",1:nsv,sep=""))
    v2 <- voom(dge,mmod1,plot=FALSE)
    next}
  
   else { an.error.occured <- FALSE}
}	
#20 minutes?
newV=NULL
fsvaobj<-fsva(v$E,mod1,svseq,newdat=NULL)
```
# Final PCA with `r nsv` factors removed

```{r,warning=FALSE,message=FALSE,fig.height = 7}
Plot_PCA(fsvaobj$db,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA with ",nsv," factors removed"))

```


```{r,warning=FALSE,message=FALSE,fig.height = 7}
contrast <- config$CONTRAST_HiSat
P_val<-config$P_VAL_HiSat
cm <- makeContrasts(contrast,levels=mmod1)
cf <- contrasts.fit(fit, cm)
fe <- eBayes(cf)  ## needed for B in topTable, but we ignore this
tt <- topTable(fe,num=Inf)
pr<-1-propTrueNull(tt$P.Value,method='convest')
cat('Conv.est.',pr,'gene level fit\n')
cfe<- eBayes(cf, proportion=pr)
adj.method='BH'


mart<-useMart(biomart="ENSEMBL_MART_ENSEMBL",
              dataset='mmusculus_gene_ensembl',
              host='uswest.ensembl.org')

geneinfo<-getBM(c("external_gene_name","ensembl_gene_id"),"ensembl_gene_id",rownames(kallisto.t.filtered),mart)
not_found<-kallisto.t.filtered[-which(rownames(kallisto.t.filtered)%in%geneinfo[,'ensembl_gene_id']),]
list_not_found <-cbind(NA,rownames(not_found))
colnames(list_not_found)<-c('external_gene_name','ensembl_gene_id')
geneinfo<-rbind(geneinfo,list_not_found)
rownames(geneinfo)<-geneinfo[,2]

res_limma<-topTable(cfe, number=Inf,
                adjust.method=adj.method, coef=colnames(cm),
                #p.value=0.05,
                #lfc=0.5,
                sort.by='none',
                confint=TRUE)
score<-apply(res_limma,1,function(x) (
    if (x[1]>0){
      x[2]}
    else
    {-x[3]}
  ))
res_limma$score<-score
res_limma$gene_name<-geneinfo[rownames(res_limma),'external_gene_name']
res_limma<-res_limma[order(res_limma$score,decreasing=TRUE),]
res_limma2<-res_limma[res_limma$adj.P.Val<P_val,]
write.csv(res_limma[,c(10,9,7,1)],paste(kallisto_path,'/Results_limma.csv',sep=''))
write.csv(res_limma2[,c(10,9,7,1)],paste(kallisto_path,'/Results_limma_pVal_',P_val,'.csv',sep=''))

```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_limma2)[1] ``

# Section 2 -EdgeR
 
```{r,warning=FALSE,message=FALSE,include=FALSE}

### edgeR
y <- DGEList(counts=kallisto.t.filtered, group=samples)
colnames(y) <- colnames(kallisto.t.filtered,)
keep <- filterByExpr(y)
summary(keep)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
design <- mmod1
y <- estimateDisp(y, design, robust=TRUE)
fit <- glmQLFit(y, design, robust=TRUE)
qlf <- glmQLFTest(fit, contrast=cm)
res_edgeR <- topTags(qlf, Inf)$table #default correction: BH
res_edgeR2<- res_edgeR[res_edgeR$PValue<P_val,]
write.csv(res_edgeR,paste(kallisto_path,'/Results_edgeR.csv',sep=''))
write.csv(res_edgeR2,paste(kallisto_path,'/Results_edgeR_pVal_',P_val,'.csv',sep=''))
#summary(decideTests(qlf))
```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_edgeR2)[1] ``


# Section 3 -DeSeq
 
```{r,warning=FALSE,message=FALSE,include=FALSE}

##deseq2
coldata <- data.frame(condition=samples, type=rep('single-read', length(samples)))
dds <- DESeqDataSetFromMatrix(countData = round(kallisto.t.filtered),
                              colData = coldata,
                              design= mmod1)
dds <- DESeq(dds)
tt <- results(dds, contrast=cm)
res_DeSeq<- tt[is.finite(tt$adj.P.Val),]
res_DeSeq2<-res_DeSeq[res_DeSeq$adj.P.Val<P_val,]
write.csv(res_DeSeq,paste(kallisto_path,'/Results_DeSeq.csv',sep=''))
write.csv(res_DeSeq,paste(kallisto_path,'/Results_DeSeq_pVal_',P_val,'.csv',sep=''))
# #####topGO
# 
# 
# # topDiffGenes <- function (allGene, thr=0.05)
# # {
# #   return(allGene < thr);
# # }
# # 
# # 
# #   for (name in colnames(cm))
# #   {
# #     for (ont  in c('BP','MF','CC'))
# #     {
# #     geneList_WTP<-get(paste(name))$adj.P.Val
# #     names(geneList_WTP)<-rownames(get(paste(name)))
# #     WTP.BP <- new("topGOdata",
# #                   description = paste('Przew',name,'_',ont,sep=''), ontology = ont,
# #                   allGenes =geneList_WTP, geneSel = topDiffGenes,
# #                   nodeSize = 10,
# #                   annot = annFUN.org, mapping="org.Mm.eg.db",ID="ensembl")
# #     
# #     #WTP.resultFisher <- runTest(WTP.BP, algorithm = "classic", statistic = "fisher")
# #    # WTP.resultFisher.e <- runTest(WTP.BP, algorithm = "elim", statistic = "fisher")
# #     WTP.resultFisher.pc <- runTest(WTP.BP, algorithm = "parentchild", statistic = "fisher")
# #     
# #     used<-usedGO(WTP.BP)
# #     WTP.allRes <- GenTable(WTP.BP, #classicFisher = WTP.resultFisher,
# #                            parentchildFisher=WTP.resultFisher.pc,orderBy = "parentchildFisher",
# #                            ranksOf = "parentchildFisher",topNodes=length(used))
# #     
# #     assign(paste(name,'.GOs.',ont,sep=''),WTP.allRes)
# #  } } 
# # 
# 
```

# Number of differentially expressed genes with adjusted p-valye < `r P_val`: `r dim(res_DeSeq2)[1] ``
