---
title: "Neuropathic pain in mouse analysis"
output: pdf_document
params:
  inpath: "../configs"
---

```{r "setup", include=FALSE}
require("knitr")
path<-params$inpath
opts_knit$set(root.dir = path)
knitr::opts_chunk$set(echo=FALSE)

```

```{r,warning=FALSE,message=FALSE,include=FALSE}

packages<-c("knitr","RColorBrewer","VennDiagram","veccompare","ggplot2","gplots","yaml","gridExtra")
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE,repos = "http://cran.us.r-project.org")
      library(x, character.only = TRUE)
    }
  }
)
if (!require("BiocManager", quietly = TRUE))
   { install.packages("BiocManager",repos = "http://cran.us.r-project.org")
BiocManager::install()}
listOfBiocPackages<-c("topGO","org.Mm.eg.db","biomaRt","limma","DESeq2","edgeR","sva")
notInstalled <- which( !listOfBiocPackages %in% rownames(installed.packages()) )
## check there's still something left to install
if( length(notInstalled) ) {
    BiocManager::install(listOfBiocPackages[ notInstalled ])
}
path_configs<-params$inpath

#opts_knit$set(root.dir = path)
knitr::opts_chunk$set(echo=FALSE)

```

```{r,warning=FALSE,message=FALSE,include=FALSE}
###########
#Functions
###########
Plot_PCA<-function(data,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,title){
    symbols<-c()
    for (i in 1:no_of_batches){
        symbols<-c(symbols,c(rep(i,each=size_of_batches[i])))
        }

    pc<-prcomp(t(data))
    pc1<-round((pc$sdev[1])^2 / sum(pc$sdev^2),digits=2)
    pc2<-round((pc$sdev[2])^2 / sum(pc$sdev^2),digits=2) 
    pc3<-round((pc$sdev[3])^2 / sum(pc$sdev^2),digits=2) 
    cols<-rep(sample(col_vector,length(names_col)),each=group_size)

    WTP<-grep("(B[1-3]_WTP_SHAM)",samples)
    PNSL<-grep("(B[1-3]_WTP_PNSL)",samples)
    cols[WTP]<-cols[21]
    cols[PNSL]<-cols[17]
    symbols2<<-symbols
    cols2<<-cols
    plot( pc$x[,1:2],pch=symbols, cex=3,lwd=4,col=cols,main=title,xlab=paste('PC1=',pc1),ylab=paste('PC2=',pc2))
    legend("topright", inset=c(-0.3,0),legend=names_col,pch=rep(unique(symbols),size_of_batches/group_size),col=unique(cols),cex=0.7)

    }

Plot_violin<-function(data,samples, group_size){

    variable<-group_size-1
    for (i in seq(1,length(samples)-variable, by=group_size))
    {
        disp<-paste(samples[i],'_disp_raw',sep="")
        disp_v<-apply(data[,i:(i+variable)],1,function(x) (
            if (sum(x)>0){
                (sd(x, na.rm=TRUE)/mean(x, na.rm=TRUE))*100}
            else{200}
        ))
        assign(paste(disp),disp_v)
    }

    df_dispersion_raw<-NULL
    for (i in (1:length(names_col)))
    {
        df_dispersion_raw<-cbind(df_dispersion_raw,get(paste(names_col[i],'_disp_raw',sep='')))
    }
    colnames(df_dispersion_raw)<-paste(names_col,'_disp_raw',sep='')
    df_dispersion_raw<-as.data.frame(df_dispersion_raw)

    pr_below_10_raw<-c()
    pr_below_20_raw<-c()
    for (i in seq(1,length(df_dispersion_raw)))
    {
        pr_below_10_raw[i]<-(length(which(df_dispersion_raw[,i]<10))/dim(df_dispersion_raw)[1])*100
        pr_below_20_raw[i]<-(length(which(df_dispersion_raw[,i]<20))/dim(df_dispersion_raw)[1])*100
    }

    df.m <- reshape2::melt(df_dispersion_raw, id.vars = NULL)
    pl<-ggplot(df.m, aes(x = variable, y = value)) + geom_violin()
    x11()
    pl + theme(axis.text.x = element_text(face="bold", color="#993333",size=10, angle=90))
}

###################
#Data load
###################
## HiSat2+StringTie
hisat.t <- read.table(file="/Users/agatamuszynska/Work/Expression_analysis_GRCm39/transcript_count_matrix.csv", header=TRUE, row.names=1, sep=",")
hisat.g <- read.table(file="/Users/agatamuszynska/Work/Expression_analysis_GRCm39/gene_count_matrix.csv", header=TRUE, row.names=1, sep=",")
Description<-read.table("/Users/agatamuszynska/Work/Pain_final/Pain_HiSat/Description.txt",sep=',',header=TRUE, stringsAsFactors=FALSE)

for (i in (1:88))
{
  Description[i,7]<-paste('B',Description[i,5],'_',Description[i,3],'_',Description[i,4],sep="")
}
colnames(Description)[7]<-c("Name")
colnam2<-c()
for (i in (1:88))
{
  colnam2[i]<-Description[which(Description$Samples==colnames(hisat.g)[i]),7]
}
colnames(hisat.t) <- colnam2
colnames(hisat.g) <- colnam2
hisat.g<-hisat.g[,order(colnames(hisat.g))]
hisat.t<-hisat.t[,order(colnames(hisat.t))]

###################
#Data description
###################

no_of_batches<-length(unique(Description$Batch))
size_of_batches<-c()
for (i in (1:no_of_batches))
{
  size_of_batches<-cbind(size_of_batches,length(which(Description$Batch==i)))
}
if(sum(size_of_batches)!=dim(hisat.g)[2]){
    stop('Sum of all batches sizes does not match input size')
    }
samples<-sort(paste('B',Description$Batch,'_',Description$Condition,sep=''))
names_col<-unique(samples)
group_size<-length(samples)/length(names_col)
#mainDir<-getwd()
#dir.create("Results")
#setwd(file.path(mainDir, "Results"))
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

######################
#Low values removal
######################

th1 <- 5
th2 <- 2
n1 <- round(dim(hisat.g)[2]/3)
n2 <- n1*2
filter1 <- apply(hisat.g, 1, function(x) length(x[x>th1])>=n1)
filter2 <- apply(hisat.g, 1, function(x) length(x[x>th2])>=n2)
filter <- filter1&filter2

hisat.g.filtered<-hisat.g[filter,]
hisat.g.filtered.log<-log2(hisat.g.filtered+0.5)

samples_PNSL_SHAM<-c()
for (i in c(1:88))
{
  if (length(grep('PNSL',samples[i]))==1)
    samples_PNSL_SHAM[i]<-'PNSL'
  else
    samples_PNSL_SHAM[i]<-'SHAM'
}

names_col2<-unique(samples_PNSL_SHAM)


 
```

# Heatmap for wildtype samlpes, raw data

```{r,warning=FALSE,message=FALSE,fig.align='left'}

heatmap.2(hisat.g.filtered.log[topVar,samples_new],
          scale="row",
          hclust=function(x) hclust(x,method="average"),
          distfun=function(x) as.dist((1-cor(t(x)))/2),
          trace="none",
          main= 'Heatmap for wildtype samples\n raw data',
          density="none",
          labRow="",
          cexCol=0.5,dendrogram='column',key=FALSE)
#dev.copy2pdf(device=dev.cur(),file="Heatmap raw data.pdf",out.type='pdf')

```

# PCA for raw data

```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=7,fig.width=8}

############
#PCA before
############
Plot_PCA(hisat.g.filtered.log,samples,names_col,group_size,no_of_batches,size_of_batches,col_vector,"PCA before SVA")
legend('topleft',legend=names_col,pch=rep(unique(symbols2),lwd=4,size_of_batches/group_size),col=unique(cols2),cex=0.7)


```


```{r,warning=FALSE,message=FALSE,include=FALSE}

########
##limma
########
# dge <- DGEList(counts=hisat.g.filtered, group=samples)
# dge <- calcNormFactors(dge)
## voom transformation
# mod1 <- model.matrix(~0+as.factor(samples))
# mod0 <- model.matrix(~1,data=as.factor(samples))
# colnames(mod1) <- levels(factor(samples))
# rownames(mod1) <- colnames(hisat.g.filtered)
# v <- voom(dge,mod1,plot=FALSE)
# to_sva <- 2^v$E
# num.sv(as.matrix(hisat.g.filtered),mod1,method="leek") 
# nsv <-25 #66->25
# svseq <- svaseq(to_sva,mod1,mod0,const=0,n.sv=nsv)
# newV=NULL
# fsvaobj<-fsva(v$E,mod1,svseq,newdat=NULL)

############################################################
#Change WTP_SHAM and PNSL for the same group in all batches
############################################################

samples_new<-sub("(B[1-3]_WTP_SHAM)","WTP_SHAM",samples)
samples_new<-sub("(B[1-3]_WTP_PNSL)","WTP_PNSL",samples_new)
dge <- DGEList(counts=hisat.g.filtered, group=samples_new)
dge <- calcNormFactors(dge)
## voom transformation
mod1 <- model.matrix(~0+as.factor(samples_new))
mod0 <- model.matrix(~1,data=as.factor(samples_new))
colnames(mod1) <- levels(factor(samples_new))
rownames(mod1) <- colnames(hisat.g.filtered)
v <- voom(dge,mod1,plot=FALSE)
to_sva <- 2^v$E
num.sv(as.matrix(hisat.g.filtered),mod1,method="leek") 


nsv <-64 #70->64->30
svseq <- svaseq(to_sva,mod1,mod0,const=0,n.sv=nsv)
newV=NULL
fsvaobj<-fsva(v$E,mod1,svseq,newdat=NULL)

names_col_new<-names_col
names_col_new<-sub("(B[1-3]_WTP_SHAM)","WTP_SHAM",names_col_new)
names_col_new<-sub("(B[1-3]_WTP_PNSL)","WTP_PNSL",names_col_new)
```



```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=7,include=FALSE}

##########
#PCA after
##########
Plot_PCA(fsvaobj$db,samples,names_col_new,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA after SVA, n.sv=",nsv))
cols3<-unique(cols2)
cols3<-c(cols3[1:12],cols3[5:6],cols3[13:18],cols3[5:6])
legend("bottomright",legend=names_col,pch=rep(unique(symbols2),size_of_batches/group_size),col=cols3,cex=0.7)
```
```{r,warning=FALSE,message=FALSE,include=FALSE}

nsv <-5 #70->64->30
svseq <- svaseq(to_sva,mod1,mod0,const=0,n.sv=nsv)
newV=NULL
fsvaobj<-fsva(v$E,mod1,svseq,newdat=NULL)

names_col_new<-names_col
names_col_new<-sub("(B[1-3]_WTP_SHAM)","WTP_SHAM",names_col_new)
names_col_new<-sub("(B[1-3]_WTP_PNSL)","WTP_PNSL",names_col_new)
```

# PCA after  svaseq with 5 factors removed

```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=7,fig.width=8}

##########
#PCA after
##########
Plot_PCA(fsvaobj$db,samples,names_col_new,group_size,no_of_batches,size_of_batches,col_vector,paste("PCA after SVA, n.sv=",nsv))
cols3<-unique(cols2)
cols3<-c(cols3[1:12],cols3[5:6],cols3[13:18],cols3[5:6])
legend("bottomleft",legend=names_col,pch=rep(unique(symbols2),size_of_batches/group_size),col=cols3,cex=0.7)
```

```{r,warning=FALSE,message=FALSE,include=FALSE}

mod1_2 <- model.matrix(~0+as.factor(samples))
mod0_2 <- model.matrix(~1,data=as.factor(samples))
colnames(mod1_2) <- levels(factor(samples))
rownames(mod1_2) <- colnames(hisat.g.filtered)
rm(mmod1)
mmod1 <- cbind(mod1_2, svseq$sv)
colnames(mmod1) <- c(levels(factor(samples)), paste("X",1:nsv,sep=""))
v2 <- voom(dge,mmod1,plot=FALSE)
#fit = lmFit(v2,mmod1,method='robust', maxit=9999)

cm <- makeContrasts(#B1_CMV_PNSLvsB1_CMV_SHAM=B1_CMV_SHAM-B1_WTP_SHAM, # mutacja CMV
                    #B1_DLX_PNSLvsB1_DLX_SHAM=B1_DLX_SHAM-B1_WTP_SHAM, #mutacja DLX
                    B1_WTP_PNSLvsB1_WTP_SHAM=B1_WTP_PNSL-B1_WTP_SHAM, #uszkodzenie
                    B1_CMV_UxM=B1_CMV_PNSL-B1_CMV_SHAM-B1_WTP_PNSL+B1_WTP_SHAM, #uszkodzeniexmutacja
                    #B1_DLX_UxM=B1_DLX_PNSL-B1_DLX_SHAM-B1_WTP_PNSL+B1_WTP_SHAM, #uszkodzeniexmutacja
                    
                    #B2_CMV_PNSLvsB2_CMV_SHAM=B2_CMV_SHAM-B2_WTP_SHAM, #mutacja CMV
                    #B2_DLX_PNSLvsB2_DLX_SHAM=B2_DLX_SHAM-B2_WTP_SHAM, #mutacja DLX
                    #B2_NAV_PNSLvsB2_NAV_SHAM=B2_NAV_SHAM-B2_WTP_SHAM, #mutacja NAV
                    B2_WTP_PNSLvsB2_WTP_SHAM=B2_WTP_PNSL-B2_WTP_SHAM, #uszkodzenie
                    B2_CMV_UxM=B2_CMV_PNSL-B2_CMV_SHAM-B2_WTP_PNSL+B2_WTP_SHAM, #uszkodzeniexmutacja
                    #B2_DLX_UxM=B2_DLX_PNSL-B2_DLX_SHAM-B2_WTP_PNSL+B2_WTP_SHAM, #uszkodzeniexmutacja
                    #B2_NAV_UxM=B2_NAV_PNSL-B2_NAV_SHAM-B2_WTP_PNSL+B2_WTP_SHAM, #uszkodzeniexmutacja
                    
                    #B3_CMV_PNSLvsB3_CMV_SHAM=B3_CMV_SHAM-B3_WTP_SHAM, #mutacja CMV
                    #B3_DLX_PNSLvsB3_DLX_SHAM=B3_DLX_SHAM-B3_WTP_SHAM, #mutacja DLX
                    #B3_NAV_PNSLvsB3_NAV_SHAM=B3_NAV_SHAM-B3_WTP_SHAM, #mutacja NAV
                    B3_WTP_PNSLvsB3_WTP_SHAM=B3_WTP_PNSL-B3_WTP_SHAM, #uszkodzenie
                    B3_CMV_UxM=B3_CMV_PNSL-B3_CMV_SHAM-B3_WTP_PNSL+B3_WTP_SHAM, #uszkodzeniexmutacja
                    #B3_DLX_UxM=B3_DLX_PNSL-B3_DLX_SHAM-B3_WTP_PNSL+B3_WTP_SHAM, #uszkodzeniexmutacja
                    #B3_NAV_UxM=B3_NAV_PNSL-B3_NAV_SHAM-B3_WTP_PNSL+B3_WTP_SHAM, #uszkodzeniexmutacja
                    
                    levels=mmod1)
cf <- contrasts.fit(fit, cm)
fe <- eBayes(cf);  ## needed for B in topTable, but we ignore this
tt <- topTable(fe,num=Inf);
pr<-1-propTrueNull(tt$P.Value,method='convest');
cat('Conv.est.',pr,'gene level fit\n'); #0.99
cfe<- eBayes(cf, proportion=pr);
adj.method='BH';
# 
# mart<-useMart(biomart="ENSEMBL_MART_ENSEMBL",
#               dataset='mmusculus_gene_ensembl',
#               host='uswest.ensembl.org',
#               ensemblRedirect=FALSE)
# 
# geneinfo<-getBM(c("external_gene_name","ensembl_gene_id"),"ensembl_gene_id",rownames(hisat.g.filtered),mart)
# not_found<-hisat.g.filtered[-which(rownames(hisat.g.filtered)%in%geneinfo[,'ensembl_gene_id']),]
# list_not_found <-cbind(NA,rownames(not_found))
# colnames(list_not_found)<-c('external_gene_name','ensembl_gene_id')
# geneinfo<-rbind(geneinfo,list_not_found)
# rownames(geneinfo)<-geneinfo[,2]

for (name in colnames(cm))
{
  res<-topTable(cfe, number=Inf,
                adjust.method=adj.method, coef=name,
                #p.value=0.05,
                #lfc=0.5,
                sort.by='none',
                confint=TRUE)
  score<-apply(res,1,function(x) (
    if (x[1]>0){
      x[2]}
    else
    {-x[3]}  
  ))
  res$score<-score
  res$gene_name<-geneinfo[rownames(res),'external_gene_name']
  res<-res[order(res$score,decreasing=TRUE),]
  res2<-res[res$adj.P.Val<0.05,]
  assign(paste(name),res)
  assign(paste(name,'.pval',sep=''),res2)
  write.csv(res2[,c(10,9,7,1)],paste(name,'.csv',sep=''))
  
  Gms<-res2[grep('Gm.*',res2$gene_name),]
  Mirs<-res2[grep('Mir.*',res2$gene_name),]
  Rest<-res2[!(rownames(res2))%in%c(rownames(Gms),rownames(Mirs)),]
  assign(paste(name,'_coding',sep=''),Rest)
  write.csv(Rest[,c(10,9,7,1)],paste(name,'_coding.csv',sep=''))
  
}
```

```{r,warning=FALSE,message=FALSE,include=FALSE}

unikatowe<-unique(c(rownames(B1_WTP_PNSLvsB1_WTP_SHAM.pval),
                    rownames(B2_WTP_PNSLvsB2_WTP_SHAM.pval),
                    rownames(B3_WTP_PNSLvsB3_WTP_SHAM.pval)))
inter<-intersect(rownames(B1_WTP_PNSLvsB1_WTP_SHAM.pval),
                    intersect(rownames(B2_WTP_PNSLvsB2_WTP_SHAM.pval),
                    rownames(B3_WTP_PNSLvsB3_WTP_SHAM.pval)))
samples_new<-grep("(B[1-3]_WTP_SHAM)",samples)
samples_new<-c(samples_new,grep("(B[1-3]_WTP_PNSL)",samples))
IQR<- apply(fsvaobj$db[,samples_new],1,IQR,na.rm=TRUE)
topVar<-order(IQR, decreasing=TRUE)[1:500]
```

# Heatmap after svaseq

```{r,warning=FALSE,message=FALSE,fig.align='left',fig.height=4}

############
#heatmap??
############


heatmap.2(as.matrix(fsvaobj$db[topVar,samples_new]),
          scale="row",
          hclust=function(x) hclust(x,method="average"),
          distfun=function(x) as.dist((1-cor(t(x)))/2),
          trace="none",
          main= 'Heatmap after svaseq',
          density="none",
          labRow="",
          cexCol=0.5,dendrogram='column',key=FALSE)
#dev.copy2pdf(device=dev.cur(),file="Heatmap after norm 1000.pdf",out.type='pdf')
```
```{r,warning=FALSE,message=FALSE,include=FALSE}

#####KEGG
#####topGO


# topDiffGenes <- function (allGene, thr=0.05)
# {
#   return(allGene < thr);
# }
# 
# 
#   for (name in colnames(cm))
#   {
#     for (ont  in c('BP','MF','CC'))
#     {
#     geneList_WTP<-get(paste(name))$adj.P.Val
#     names(geneList_WTP)<-rownames(get(paste(name)))
#     WTP.BP <- new("topGOdata",
#                   description = paste('Przew',name,'_',ont,sep=''), ontology = ont,
#                   allGenes =geneList_WTP, geneSel = topDiffGenes,
#                   nodeSize = 10,
#                   annot = annFUN.org, mapping="org.Mm.eg.db",ID="ensembl")
#     
#     #WTP.resultFisher <- runTest(WTP.BP, algorithm = "classic", statistic = "fisher")
#    # WTP.resultFisher.e <- runTest(WTP.BP, algorithm = "elim", statistic = "fisher")
#     WTP.resultFisher.pc <- runTest(WTP.BP, algorithm = "parentchild", statistic = "fisher")
#     
#     used<-usedGO(WTP.BP)
#     WTP.allRes <- GenTable(WTP.BP, #classicFisher = WTP.resultFisher,
#                            parentchildFisher=WTP.resultFisher.pc,orderBy = "parentchildFisher",
#                            ranksOf = "parentchildFisher",topNodes=length(used))
#     
#     assign(paste(name,'.GOs.',ont,sep=''),WTP.allRes)
#  } } 
# 
# write.csv(B1_CMV_UxM.GOs.BP[1:100,],'B1_CMV_UxM.GOs.BP.csv')
# write.csv(B2_CMV_UxM.GOs.BP[1:100,],'B2_CMV_UxM.GOs.BP.csv')
# write.csv(B3_CMV_UxM.GOs.BP[1:100,],'B3_CMV_UxM.GOs.BP.csv')
# write.csv(B1_CMV_UxM.GOs.MF[1:100,],'B1_CMV_UxM.GOs.MF.csv')
# write.csv(B2_CMV_UxM.GOs.MF[1:100,],'B2_CMV_UxM.GOs.MF.csv')
# write.csv(B3_CMV_UxM.GOs.MF[1:100,],'B3_CMV_UxM.GOs.MF.csv')
# write.csv(B1_CMV_UxM.GOs.CC[1:100,],'B1_CMV_UxM.GOs.CC.csv')
# write.csv(B2_CMV_UxM.GOs.CC[1:100,],'B2_CMV_UxM.GOs.CC.csv')
# write.csv(B3_CMV_UxM.GOs.CC[1:100,],'B3_CMV_UxM.GOs.CC.csv')
# #####KEGG
#pho_KEGGresult<-find_enriched_pathway(res2,species='mmu')

#egIDs <- stack(mget(rownames(hisat.g.filtered), org.Mm.egENSEMBL2EG, ifnotfound = NA))
save.image('Data_all.RData')
